\\Large The heat is on: Rising temperatures alter how and where mammals move

### 2025-09-02 writing/manuscript.Rmd
updated first paragraph and table of results

In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement behaviour during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017]. We discuss the consequences of these changes in movement behaviour at the individual, population, and community levels. We also provide directions for future research and solutions for overcoming some common data limitations.

@@

\noindent Overall, 2.6% of GPS locations had temperatures lower than $-20^\circ$C, while 6.5% had temperatures above $20^\circ$C, and temperature ranges differed across species (Table \@ref(tab:temp-qs), Fig. B2). At $0^\circ$C, species differed in overall mean probabilities of movement (range: 0.05 -- 0.31), mean speed when moving (range: 0.42 -- 2.67 km/h), and mean distance traveled (i.e., $P(M) \times \mathbb E(S)$, range: 0.04 -- 0.61 km/h; Table \@ref(tab:gam-intercepts)). Grizzly bears tended to move least often ($P(M) \approx 0.05$), while wolves and cougars moved most often ($P(M) \ge 0.22$). When moving, mountain goats and southern mountain caribou moved the slowest ($\mathbb E(S|M) \approx 0.43$ km/h), while wolves had the highest mean speed when moving ($\mathbb E(S|M) \approx 2.64$ km/h). Consequently, wolves traveled, on average, 0.6 km/h -- 2.5 to 16.7 times further than other mammals at $0^\circ$C.

```{r temp-qs}
library('dplyr')
d <- readRDS('../data/hgam-speed-data.rds')
get_perc <- function(bools) {
  out <- as.character(round(mean(bools) * 100, 1))
  if(grepl('\\.', out)) {
    out <- paste0(out, ' %')
  } else {
    out <- paste0(out, '.0 %')
  }
  return(out)
}

bind_rows(
  d %>%
    rename(Species = species) %>%
    group_by(Species) %>%
    summarize(`T < -20\\textdegree{} C` = get_perc(temp_c < -20),
              `T > +20\\textdegree{} C` = get_perc(temp_c > +20)),
  d %>%
    summarize(Species = 'Total',
              `T < -20\\textdegree{} C` = get_perc(temp_c < -20),
              `T > +20\\textdegree{} C` = get_perc(temp_c > +20))) %>%
  knitr::kable(format = 'markdown', caption = paste0("Percentage of observed GPS locations (after data cleaning) with temperatures below -20\U00B0", "C or above 20\U00B0", "C."))
```

\scriptsize

```{r gam-intercepts}
setwd('~/GitHub/bc-mammals-temperature')
library('mgcv')
library('dplyr')
library('ctmm') #' for `%#%`
source('analysis/figures/default-ggplot-theme.R')

@@

intercepts <- tibble(
  Species = SPECIES,
  'P(M)' = get_int(readRDS('models/binomial-gam-without-temperature.rds')),
  'P(M|T)' = get_int(readRDS('models/binomial-gam.rds')),
  'E(S)' = 'km/h' %#% get_int(readRDS('models/gamma-gam-without-temperature.rds')),
  'E(S|T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
  'E(D)' = `P(M)` * (`E(S)`),
  'E(D|T)' = `P(M|T)` * (`E(S|T)`)) %>%
  mutate(Species = case_when(
    Species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus (sm)',
    Species == 'Ursus arctos horribilis' ~ 'U. arctos horr.',
    Species == 'Puma concolor' ~ 'Puma concolor',
    Species == 'Cervus canadensis' ~ 'C. canadensis',
    Species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus (b)',
    Species == 'Canis lupus' ~ 'Canis lupus',
    Species == 'Oreamnos americanus' ~ 'O. americanus')) %>%
  mutate(Species = paste0('\\emph{', Species, '}') %>%
           gsub(')}', ')', .) %>%
           gsub(' \\(', '} \\(', .))

if(FALSE) {
  hist(intercepts$`P(M)`)
  hist(intercepts$`E(S)`)
  round(range(max(intercepts$`E(D)`) / intercepts$`E(D)`[- which.max(intercepts$`E(D)`)]), 1)

  intercepts %>%
    transmute(Species,
              p = round(`P(M|T)` - `P(M)`, 2),
              s = round(`E(S|T)` - `E(S)`, 2),
              d = round(`E(D|T)` - `E(D)`, 2))

  intercepts %>%
    arrange(desc(`E(D)`)) %>%
    mutate(displ_wolf = `E(D)`[grepl('Canis lupus', Species)] / `E(D)`,
           displ_wolf_t = `E(D|T)`[grepl('Canis lupus', Species)] / `E(D|T)`)
}

intercepts %>%
  arrange(Species) %>%
  rename_with(\(.x) paste0('$', .x, '$'), ! Species) %>%
  rename_with(\(.x) gsub('\\|', '\\\\vert ', .x)) %>%
  rename_with(\(.x) gsub('P', '\\\\hat{P}', .x)) %>%
  rename_with(\(.x) gsub('E', '\\\\hat{\\\\mathbb E}', .x)) %>%
  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Mean probability of movement (P(M)), speed ($\\mathbb E(S)$, km/h, assuming constant movement), and distance travelled ($P(M) \\times \\mathbb E(S) = \\mathbb E(D)$, km/h) as estimated by models without and with temperature ($\\mathbb E(... | T)$), after post-stratifting to a 1-hour sampling rate and a temperature of T = 0\U00B0", "C."))
```

>>>>

In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of moving, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement behaviour during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017]. We discuss the consequences of these changes in movement behaviour at the individual, population, and community levels. We also provide directions for future research and solutions for overcoming some common data limitations.

@@

\noindent Overall, 2.6% of GPS locations had temperatures lower than $-20^\circ$C, while 6.5% had temperatures above $20^\circ$C, and temperature ranges differed across species (Table \@ref(tab:hgam-summary), Fig. B2). Sampling interval affected estimates of probability as well of as speed when moving (Fig. B8). Consequently, we present all results while predicting specifically for one-hour sampling intervals. At $0^\circ$C, species differed in mean probabilities of moving ($\hat P(M)$; range: 0.05 -- 0.31), mean speed when moving ($\hat {\mathbb E}(S|M=1)$; range: 0.42 -- 2.67 km/h), and mean distance traveled (i.e., $\hat P(M) \times \hat {\mathbb E}(S|M=1)$, range: 0.04 -- 0.61 km/h; Table \@ref(tab:hgam-summary)). Grizzly bears had the lowest movement frequency ($\hat P(M) \approx 0.05$), while wolves and cougars moved most often ($\hat P(M) \ge 0.22$). Mountain goats and southern mountain caribou moved the slowest ($\hat{\mathbb E}(S|M = 1) \approx 0.43$ km/h), while wolves had the highest mean speed when moving ($\hat{\mathbb E}(S|M = 1) \approx 2.67$ km/h). Consequently, at $0^\circ$C, wolves traveled an average of $0.22 \times 2.67~\text{km/h} \approx 0.6$ km/h; 2.5 to 16.7 times further than other species.

\tiny

```{r hgam-summary}
setwd('~/GitHub/bc-mammals-temperature')
library('mgcv')
library('dplyr')
library('ctmm') #' for `%#%`
source('analysis/figures/default-ggplot-theme.R')

d <- readRDS('data/hgam-speed-data.rds')
get_perc <- function(bools) {
  perc <- round(mean(bools) * 100, 1)
  out <- as.character(perc)
  if(grepl('\\.', out)) {
    out <- paste0(out, ' %')
  } else {
    out <- paste0(out, '.0 %')
  }

  if(perc < 10) out <- paste0(' ', out)

  return(out)
}

@@

intercepts <- tibble(
  Species = SPECIES,
  'P(M=1)' = get_int(readRDS('models/binomial-gam-without-temperature.rds')),
  'P(M=1|T)' = get_int(readRDS('models/binomial-gam.rds')),
  'E(S)' = 'km/h' %#% get_int(readRDS('models/gamma-gam-without-temperature.rds')),
  'E(S|T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
  'E(D)' = `P(M=1)` * (`E(S)`),
  'E(D|T)' = `P(M=1|T)` * (`E(S|T)`)) %>%
  mutate(Species = case_when(
    Species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus; sm',
    Species == 'Ursus arctos horribilis' ~ 'U. arctos h.',
    Species == 'Puma concolor' ~ 'P. concolor',
    Species == 'Cervus canadensis' ~ 'C. canadensis',
    Species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus; b',
    Species == 'Canis lupus' ~ 'Canis lupus',
    Species == 'Oreamnos americanus' ~ 'O. americanus')) %>%
  mutate(Species = paste0('\\emph{', Species, '}') %>%
           gsub(')}', ')', .) %>%
           gsub(' \\(', '} \\(', .))

if(FALSE) {
  # visualize the individual estimates
  layout(1:3)
  hist(intercepts$`P(M=1)`); hist(intercepts$`E(S)`); hist(intercepts$`E(D)`)
  layout(1)

  # find differences when accounting for temperature
  intercepts %>%
    transmute(Species,
              p = round(`P(M=1|T)` - `P(M=1)`, 2),
              s = round(`E(S|T)` - `E(S)`, 2),
              d = round(`E(D|T)` - `E(D)`, 2))

  # find difference in distance traveled, relative to furtherst distance
  intercepts %>%
    arrange(desc(`E(D)`)) %>%
    mutate(displ_wolf = `E(D)`[grepl('Canis lupus', Species)] / `E(D)`,
           displ_wolf_t = `E(D|T)`[grepl('Canis lupus', Species)] / `E(D|T)`)

  # find difference in distance traveled, relative to furtherst distance
  round(range(max(intercepts$`E(D|T)`) / intercepts$`E(D|T)`[- which.max(intercepts$`E(D|T)`)]), 1)
}

options(knitr.kable.NA = '') # leave NAs blank

bind_rows(
  # species-level proportions
  d %>%
    rename(Species = species) %>%
    group_by(Species) %>%
    summarize(`T<-20\\textdegree{}C` = get_perc(temp_c < -20),
              `T>+20\\textdegree{}C` = get_perc(temp_c > +20)),
  # overall proportions
  d %>%
    summarize(Species = 'Total',
              `T<-20\\textdegree{}C` = get_perc(temp_c < -20),
              `T>+20\\textdegree{}C` = get_perc(temp_c > +20))) %>%
  select(! Species) %>%
  bind_cols(intercepts %>%
              arrange(Species) %>%
              rename_with(\(.x) paste0('$', .x, '$'), ! Species) %>%
              rename_with(\(.x) gsub('\\|', '\\\\vert ', .x)) %>%
              rename_with(\(.x) gsub('P', '\\\\hat{P}', .x)) %>%
              rename_with(\(.x) gsub('E', '\\\\hat{\\\\mathbb E}', .x)) %>%
              bind_rows(tibble(Species = 'Total'))) %>%
  relocate(Species, .before = 1) %>%
  rename(`$\\hat{P}$(M = 1)` = `$\\hat{P}(M=1)$`) %>% # to avoid eq going to new line
  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Summary statistics for the observed temperatures and estimated mean movement rates at \U00B0", "C. The second and third columns indicate the percentage of GPS locations (after data cleaning) with temperature ($T$) below -20\U00B0", "C and above 20\U00B0", "C, respectively. The subsequent columns indicate the estimated mean probability of moving ($\\hat P(M=1)$), mean speed when moving ($\\hat{\\mathbb E}(S)$; km/h), and mean distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S) = \\hat{\\mathbb E}(D)$; km/h) as estimated by models without and with temperature (i.e., $(... | T)$), for a sampling interval of 1 hour and a temperature of T = 0\U00B0", "C."))
```
